
```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

[![Build Status](https://travis-ci.org/mpadge/paretoconv.svg)](https://travis-ci.org/mpadge/paretoconv) 
[![Project Status: Concept - Minimal or no implementation has been done yet.](http://www.repostatus.org/badges/0.1.0/concept.svg)](http://www.repostatus.org/#concept)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/paretoconv)](http://cran.r-project.org/web/packages/paretoconv) 
[![codecov](https://codecov.io/gh/mpadge/paratoconv/branch/master/graph/badge.svg)](https://codecov.io/gh/mpadge/paretoconv)

`paretoconv` contains only one function:
```
paretoconv (x, a, n)
```
which calculates the $n$-fold convolution of two Pareto distributions,
\begin{equation}
f(x)=a x^{1-a},
\end{equation}
for $(x,a)>0$ using the techniques devised by Colin Ramsay in

1. 'The Distribution of Sums of Certain I.I.D.  Pareto Variates' (Communications
   in Statistics - Theory and Methods 35:395-405, 2006); and

2. 'The Distribution of Sums of I.I.D. Pareto Random Variables with Arbitrary
   Shape Parameter' (Communications in Statistics - Theory and Methods
   37:2177-2184, 2008).


### News

- Initial working version 

### Installation

```{r, eval=FALSE}
devtools::install_github("mpadge/paretoconv")
```

```{r, echo=FALSE, message=FALSE}
wd <- getwd ()
while (length (grep ('paretoconv', getwd ())) > 0) setwd ("..")
devtools::load_all ("paretoconv")
#devtools::load_all ("paretoconv", recompile=TRUE)
devtools::document ("paretoconv")
#goodpractice::gp ("paretoconv")
setwd (wd)
#testthat::test_dir ("./tests/")
```

```{r echo=FALSE, message=FALSE, warning=FALSE, error=FALSE}
options(width=120)
```

### Usage

```{r}
library(paretoconv)
# current verison
packageVersion("paretoconv")
paretoconv (x=0:5, a=1, n=5)
```
Note that `x` can be a vector, while both `a` and `n` must single-valued.

### Test Results

```{r, eval=FALSE}
library(paretoconv)
library(testthat)

date()

test_dir("tests/")
```
