---
title: "paretoconv"
author: "Mark Padgham"
date: "`r Sys.Date()`"
output:
    html_document:
        toc: true
        toc_float: true
        theme: flatly
        number_sections: true
vignette: >
  %\VignetteIndexEntry{paretoconv}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r load-packages, message=FALSE}
devtools::load_all (".", export_all=FALSE)
library (poweRlaw)
```

The following code analyses the American civil war data included in the
`poweRlaw` package data.

-------


# Generating cumulative distribution functions

Cumulative Distribution Functions (CDFs) are not appropriately fitted with
conventional methods, because these assume an absolute lower limit, and simply
normalise to  (empirical) unit integral above that limit. This section
implements the method described in the paper to properly scale CDFs.

If the maximal distance between observed ($y_i$) and modelled ($z_i$) values is
$(z_a - y_a)$ for some particular $a$--where it is explicitly assumed that $z_a
> y_a$---and the corresponding minimal distance is $(y_b - z_b)$ for some $b$, 
then there will exist some value, $k$, such that,
\begin{equation}
    k z_a - y_a = y_b - k z_b,
\end{equation}
such that the observed values will extend both above and below the modelled
values by the same extent. This extent will obviously represent the minimal
possible value of the Kolgorov-Smirnov (KS) statistic, and will arise with a
value of,
\begin{equation}
    k = \frac {y_a + y_b} {z_a + z_b}.
\end{equation}
The value of $k$ requires initially determining the values of $a$ and $b$ at
which the respective maxima arise. The modelled values are the multiplied by
this value of $k$, and the values of $a$ and $b$ once again determined, along
with corresponding KS statisics. These subsequent values of $a$ and $b$ may
differ from the initial values, and the procedure must be repeated until the
values of $a$ and $b$--the positions at which the maxima arise--converge.



## Straight power law

The following fits a `poweRlaw` discrete power law object to the American
casualities data.

```{r displ-plot}
data ("us_american")
x <- us_american$Cas
m_us <- displ$new(x)
#m$setPars (estimate_pars (m_us))
m_us$setXmin (estimate_xmin (m_us))
a <- m_us$pars
x0 <- m_us$xmin

freq <- as.vector (table (sort (x)))
cum_n <- rev (cumsum (rev (freq)))
cum_n <- cum_n / max (cum_n)

x <- sort (unique (x))
y <- (x0 / x) ^ (a - 1)
```

The observed and modelled values are then respectively represented by `cum_n`
and `y`. The iterative procedure is implemented with the following code, which
simply stops when the value of $k$ converges to one, because the maximal
difference arises for the specified values of $\{z_i\}$ without any additional
multiplication.
```{r fit_ks_model}
fit_ks_model <- function (yobs, ymod, k = 10)
{
    n <- 0
    while (abs (k - 1) > 1e-10)
    {
        i1 <- which.min (yobs - ymod)
        i2 <- which.max (yobs - ymod)
        k <- (yobs [i1] + yobs [i2]) / (ymod [i1] + ymod [i2])
        ymod <- ymod * k
        n <- n + 1
    }
    message ('converged in ', n, ' iterations')
    if (abs (max (yobs - ymod) - max (ymod - yobs)) > 1e-10)
        message ("Iterative fitting of KS statistic did not yield equal
                 estimates")
    list (fit = ymod, ks = max (yobs - ymod))
}
ft <- fit_ks_model (cum_n, y, k = 10)
ft$ks
```

The difference between this approach and the conventional approach used in
`poweRlaw` can be seen in the following plot:
```{r plot1, fig.height=6}
plot (m_us)
lines (m_us, col="red", lwd=1)
lines (x, ft$fit, col="blue", lty=2)
```

## Convoluted power law

Follow the same procedure for the native American casualty data which reflect
the convolution of multiple power-law processes.

```{r displ-plot2}
data ("native_american")
x <- native_american$Cas
m_na <- displ$new(x)
m_na$setXmin (estimate_xmin (m_na))
a <- m_na$pars
x0 <- m_na$xmin

freq <- as.vector (table (sort (x)))
cum_n <- rev (cumsum (rev (freq)))
cum_n <- cum_n / max (cum_n)

x <- sort (unique (x))
y <- paretoconv (x, a=a, x0 = 3, n = 2, cdf=TRUE)
```

```{r fit_ks_model-demo}
ft <- fit_ks_model (cum_n, y, k = 10)
ft$ks
```

And then the actual plot
```{r, fig.height=6}
plot (m_na)
lines (m_na, col="red", lwd=1)
lines (x, ft$fit, col="blue", lty=2)
```


# Optimal `paretoconv` models

Optimal `paretoconv` models simply refer to values of $n$ and $x_0$. These have
to be found using a form of discrete optimisation, currently implemented in the
following function:
```{r pareto_optimise, eval = FALSE}
pareto_optimise (m_na, x0 = 1, n = 1, quiet = FALSE)
```
Revealing optimal values of $x_0 = 3$ and $n = 2$, with a KS distance of
`0.02782716`. What is needed is a discrete optimisation algorithm, which doens't
actually exist in any `R` package

## Discrete Optimisation

Because the maximum for most discrete cases is likely to be for fairly small
values of `n` and `x0`, the optimisation problem is best solved by brute force.
(Alternatives such as hard-coding a Newton-type algorithm are not really
practical, because they all generate non-integer step sizes which can't be
readily implemented.) A brute force gradient approach is implemented as
`pareto_optimise()`, which repeatedly examines the $\pm1$ values around $(x_0,
n)$, and discerns the values yielding the minimal KS statistics. The routine
stops on convergence, yielding the following results:
```{r optimise1, eval = FALSE}
system.time (
             x <- pareto_optimise (m_na, x0 = 1, n = 1, quiet = FALSE)
             )
x
```
```{r, echo=FALSE}
x <- c (3, 2, 0.02782716)
names (x) <- c ('x0', 'n', 'KS')
tt <- c (24.337, 0.000, 24.337)
names (tt) <- c ("user", "system", "elapsed")
message ("checking whether non-convoluted version is optimal: NO\n",
         "calculating initial KS statistics ...")
tt; x
```

And the equivalent values for US casualities:
```{r optimise2, eval = FALSE}
system.time (
             x <- pareto_optimise (m_us, x0 = 1, n = 1, quiet = FALSE)
             )
x
```
```{r, echo=FALSE}
x [1:3] <- c (1, 1, 0.05824103)
tt [1:3] <- c (6.137, 0.000, 6.138)
message ("checking whether non-convoluted version is optimal: NO\n",
         "calculating initial KS statistics ...")
tt; x
```



------------------

## Testing best model

The preceding KS distance metrics may be used to establish the best model, but
that model still needs to be statistically tested against the observed data.
This is done using synthetic data generated with `rplconv()`, which is simply
`poweRlaw::rpldis` modified to generate random deviates across the full range
including `0<x<xmin`.

```{r conv-synth-data}
a <- m_na$getPars ()
n <- 2
nx <- 1000
xt <- rep (0, nx)
set.seed (0)
for (i in 1:n)
    xt <- xt + rplconv (n = nx, x0 = 3, alpha=a)
xt <- floor (xt / n)
```
A `poweRlaw` model then has to be generated for these data by first estimating
the value of `Xmin`:
```{r conv-synth-data-plot}
m_sy <- displ$new (xt)
m_sy$setXmin (poweRlaw::estimate_xmin (m_sy))
plot (m_sy)
lines (m_sy, col="red", lwd=2)
```

Again, the `plot` method of `poweRlaw` just does this:
```{r conv-synth-data2, eval = FALSE}
x0 <- as.vector (table (sort (xt)))
y0 <- rev (cumsum (rev (x0)))
y0 <- y0 / max (y0)
plot (unique (sort (xt)), y0, pch=1, log="xy")
```
Proper testing requires values for `x0` and `n` to be estimated for each set of
simulated data
```{r optimise3, eval = FALSE}
pareto_optimise (m_sy, x0 = 3, n = 2, quiet = FALSE)
```
```{r, echo=FALSE}
junk <- c (1, 2, 0.0326965)
names (junk) <- c ('x0', 'n', 'KS')
junk
```
That demonstrates the principle of testing an observed model, as well as the
fact that synthetic models often correspond to different values of `x0` and `n`
than the original models, and thus optimal models need to be fitted to synthetic
data. This is time consuming, yet currently incorporated within the `paretoconv`
function `pparetoconv()`, which generates a number of synthetic models (values
of `a`, `x0` and `n`), uses those to generate synthetic series (using
`rplconv()`), and derives the probability of observing a more extreme
KS-statistic than the observed value.

```{r pparetoconv, eval = FALSE}
st <- Sys.time ()
pparetoconv (m_na, x0 = 3, n = 2, quiet = FALSE)
Sys.time () - st
```
```{r, echo=FALSE}
message ("Generating simulated models\n",
         "Generating models until same model appears 4 times\n\n",
         "model#01: (x0, n) = (2, 2)\n",
         "model#02: (x0, n) = (3, 2)\n",
         "model#03: (x0, n) = (3, 1)\n",
         "model#04: (x0, n) = (1, 3)\n",
         "model#05: (x0, n) = (3, 1)\n",
         "model#06: (x0, n) = (3, 1)\n\n",
         "Generating synthetic series from 3 simulated models\n",
         "mod [2, 2] ..........\n",
         "mod [3, 1] ........................................\n",
         "mod [1, 3] ..........\n",
         "Generated 60 synthetic KS statistics\n")
0.9896698
message ("Time difference of 3.910042 mins")
```
And that really does take quite a long time, yet finally confirms that the
observed KL statistic can be generated by neutral models, which suggests it is
indeed an accurate model of the underlying data.


-------

## Probability for non-convoluted model

The following code just demonstrates that the procedure works just as well for
non-convoluted models of the type analysed with `poweRlaw`.

```{r par-opt-moby, eval = FALSE}
data ('moby', package='poweRlaw')
m <- poweRlaw::displ$new (moby)
m$setXmin (poweRlaw::estimate_xmin (m))
pareto_optimise (m, x0 = 1, n = 0, quiet = FALSE)
pparetoconv (m, x0 = 1, n = 0, quiet = FALSE)
```
```{r, echo=FALSE}
message ("Generating simulated models\n",
         "Generating models until same model appears 4 times\n\n",
         "model#01: (x0, n) = (1, 0)\n",
         "model#02: (x0, n) = (1, 0)\n",
         "model#03: (x0, n) = (1, 0)\n",
         "model#04: (x0, n) = (1, 0)\n\n",
         "Generating synthetic series from 1 simulated models\n",
         "mod [1, 0] ........................................\n",
         "Generated 40 synthetic KS statistics")
0.8884125
```


------------------

# Convolutions of hetergeneous distributions

Heterogeneous distributions can be modelled using `poweRlaw::dpldis`
function (or `dplcon` for continuous distributions).

## Pareto optimisation of `poweRlaw` simulated values

Note first that values simulated with `rpldis` yield models which can be
recreated using `pareto_optimise`:

```{r Heterogeneous-conv, eval = TRUE}
library (poweRlaw)
alpha <- 3
xmin <- 1
len <- 1e4
delta <- (-1:1) / 2 # hard-coded values
x <- lapply (delta, function (i) rpldis (len, xmin = xmin, alpha = alpha + i))
x <- Reduce (`+`, x)
x <- floor (x / length (delta))
m <- displ$new(x)
#m$setPars (estimate_pars (m))
m$setXmin (estimate_xmin (m))
plot (m)
lines (m, col = "red")
slot (m, ".xData")$pars
```
This code produces realistic models, but the estimated parameters do not
reproduce the stated value of `alpha`. Wrap in a function to examine the
results:
```{r get_slope1}
get_slope1 <- function (xmin = 1, alpha = 3, delta = 0.5, len = 1e4)
{
    alpha0 <- alpha
    alpha <- c (alpha0 - delta, alpha0 + delta)
    get_plmod <- function (alpha)
    {
        x <- lapply (alpha, function (i)
                     rpldis (len, xmin = xmin, alpha = i))
        x <- floor (Reduce (`+`, x) / length (delta))
        m <- displ$new(x)
        estimate_pars (m)$pars
    }
    p1 <- get_plmod (alpha)

    alpha <- c (alpha, alpha0)
    res <- c (p1, get_plmod (c (alpha, alpha0)))
    names (res) <- c ("O2", "O3")
    return (res)
}
delta <- 1:10 / 10
s <- -do.call (rbind, lapply (delta, function (i) get_slope1(delta = i)))
plot (delta, -3 + delta, "l", col = "black", ylab = "power-law slope",
      ylim = range (c (as.vector (s), -3, -2)))
lines (delta, s [, 1], col = "red")
lines (delta, s [, 2], col = "blue")
legend ("topleft", lwd = 1, col = c ("black", "red", "blue"),
        legend = c ("theoretical", "O2", "O3"), bty = "n")
title (main = "Power law slopes for different values of delta")
```

These observed lines do not really reflect the expected values at all.
Nevertheless, fitting resultant models does converge on the expected number of
convlutions:
```{r fit-hetero-conv, eval = FALSE}
pareto_optimise (m, x0 = 1, n = 0, quiet = FALSE)
```
```{r, echo = FALSE}
message ("checking whether non-convoluted version is optimal: NO\n",
         "calculating initial KS statistics ...\n",
         "iteration#1 -> (x0, n) = (2, 1)\n",
         "iteration#2 -> (x0, n) = (3, 2)\n",
         "iteration#3 -> (x0, n) = (3, 3)\n",
         "iteration#4 -> (x0, n) = (2, 4)")
x <- c (2.0000000, 4.0000000, 0.0484632)
names (x) <- c ("x0", "n", "KS")
x
```


## Function to generate heterogeneous convoluted power law distributions

The convolution can be generated with `dpldis`, with integral from
https://stackoverflow.com/questions/23569133/adding-two-random-variables-via-convolution-in-r#23569980
CG currently removes all values below xmin from analysis here:
https://github.com/csgillespie/poweRlaw/blob/master/pkg/R/pldis.R#L28
This code has to use my current (25/01/18) PR:

```{r junk, echo = FALSE, eval = TRUE}
#devtools::load_all ("/data/mega/code/forks/poweRlaw/pkg", export_all = FALSE)
```


```{r dpldis, eval = TRUE}
f <- function(x, xmin = 1, alpha = 3)
    dpldis(x, xmin = xmin, alpha = alpha, log = FALSE)
fz0 <- Vectorize (function(z) integrate (function (x, z) f (z - x) * f (x),
                              lower = -Inf, upper = Inf, z)$value)
fz2 <- Vectorize (function(z) integrate (function (x, z)
                                         f (z - x, alpha = 2.5) * f (x, alpha = 3.5),
                              lower = -Inf, upper = Inf, z)$value)
z <- 10:1000 / 10
y0 <- f (z)
plot (z, y0, "l", log = "xy")

yz0 <- fz0 (z)
yz0 <- yz0 * min (y0) / min (yz0 [yz0 > 0])
indx <- which (yz0 != 0)
lines (z [indx], yz0 [indx], col = "blue")
yz2 <- fz2 (z)
yz2 <- yz2 * min (y0) / min (yz2 [yz2 > 0])
indx <- which (yz2 != 0)
lines (z [indx], yz2 [indx], col = "skyblue")

yp <- paretoconv (x=z, a=3, n=1, x0 = 1)
yp <- yp * min (y0) / min (yp)
lines (z, yp, col = "red")
legend ("topright", lwd = 1, col = c ("black", "blue", "skyblue", "red"),
        legend = c ("no conv", "conv (delta = 0)", "conv (delta = 0.5)",
                    "conv (Ramsay)"), bty = "n")
```

Although the heads of the `dpldis` curves do not really accord with the formal
result of Colin Ramsay, the tails do seem to converge to the expected values. In
particular:
```{r, eval = FALSE}
indx <- which (z > 20)
summary (lm (log10 (yz0 [indx]) ~ log10 (z [indx])))$coefficients [2]
summary (lm (log10 (yp [indx]) ~ log10 (z [indx])))$coefficients [2]
summary (lm (log10 (yz2 [indx]) ~ log10 (z [indx])))$coefficients [2]
```
And the latter is as expected almost precisely 0.5 less than the other estimates
of ~3. Wrap this in a function to estimate the slopes only:
```{r get_slope, eval = FALSE}
get_slope <- function (xmin = 1, alpha = 3, delta = 0.5)
{
    f <- function(x, xmin = 1, alpha = 3)
        dpldis(x, xmin = xmin, alpha = alpha, log = FALSE)
    fz <- Vectorize (function(z) integrate (function (x, z)
                                             f (z - x, alpha = alpha - delta) *
                                                 f (x, alpha = alpha + delta),
                                             lower = -Inf, upper = Inf, z)$value)
    z <- 10:1000 / 10
    y0 <- f (z)

    yz <- fz (z)
    yz <- yz * min (y0) / min (yz [yz > 0])
    indx <- which (z > 20)
    summary (lm (log10 (yz [indx]) ~ log10 (z [indx])))$coefficients [2]
}
delta <- 0:10 / 10
s <- vapply (delta, function (i) get_slope (delta = i), numeric (1))
plot (delta, -3 + delta, "l", ylab = "power-law slope")
lines (delta, s, col = "red")
legend ("topleft", lwd = 1, col = c ("black", "red"),
        legend = c ("theoretical", "observed"), bty = "n")
title (main = "Power law slopes for different values of delta")
```

The following code implements a third-order convolution, but does not run
because either `maximum number of subdivisions reached` or `roundoff error was
detected`.
```{r get_slope_O3, eval = FALSE}
xmin <- 1
alpha <- 3
delta <- 0.5
f <- function(x, xmin = 1, alpha = 3)
    dpldis(x, xmin = xmin, alpha = alpha, log = FALSE)
fz2 <- Vectorize (function(z) integrate (function (x, z)
                                         f (z - x, alpha = alpha - delta) *
                                             f (x, alpha = alpha + delta),
                                         lower = -Inf, upper = Inf, z)$value)
fz3 <- Vectorize (function (z) integrate (function (x, z)
                                          fz2 (z - x) * f (x, alpha = alpha),
                                          lower = -Inf, upper = Inf, z)$value)
z <- 10:100 / 10
z <- 1:10
yz <- fz3 (z)
```

The `O(2)` results nevertheless produce the desired demonstration that
second-order convolution of power law distributions accords precisely with the
theoretical expectations from exponential functions, in that observed slopes are
precisely biased by the value of $\Delta$.



------------------


# Comparing Distributions (old stuff)

Clauset *et. al* (2009) present a justification for using log-likehihood ratios
to compare model distributions, following Vuong (1989). Such calculations
presume that candidate models have already been demonstrated to provide
appropriate descriptions of observed data. These log-likelihood tests are thus
not the appropriate tool for comparing different forms of convoluted power-law
distributions. 

In particular, log-likelihood ratios will always be maximised for the model with
the highest mean (log-)probabilities, and without implementing an appropriate
initial selection, this will always be the model with the lowest `x0`. In the
present context, models can therefore only be compared using some objective
metric of model performance, which Clauset *et. al* convincingly argue should be
a simple maximal ('KL') distance bewteen observed and modelled distributions
(*not* log-scaled).

Although therefore not required, the following code reproduces Gillespie's
`compare_distributions` function, which requires an alternative distribution to
be generated for these data:
```{r compare_distributions}
library (poweRlaw)
data ("native_american", package="poweRlaw")
x <- native_american$Cas
m1 <- poweRlaw::displ$new(x)
m1$setPars (poweRlaw::estimate_pars (m1))
m1$setXmin (poweRlaw::estimate_xmin (m1))
m2 <- poweRlaw::disexp$new (x)
m2$setPars (poweRlaw::estimate_pars (m2))
m2$setXmin (m1$getXmin ())
plot (m1)
lines (m1, col="red", lwd=2)
lines (m2, col="orange", lwd=2, lty=2)
comp <- poweRlaw::compare_distributions (m1, m2)
message ("test_statistic = ", comp$test_statistic, " (p = ", comp$p_two_sided, ")")
```

These values are then reproduced here using a stripped down version of 
[`compare_distributions`](https://github.com/csgillespie/poweRlaw/blob/master/pkg/R/compare_distributions.R#L72-L94),
hard-coded so that `d1` is a discrete power law (`dpl`), and `d2` is an
exponential.  The function
[`dist_pdf` for a `dpl` is defined](https://github.com/csgillespie/poweRlaw/blob/master/pkg/R/pldis.R#L27-L42)
with parameters `(m, q=NULL, log=FALSE)`, which are passed to `dpldis (q[q >=
m$xmin], m$xmin, m$pars, TRUE)`. The equivalent definition for an exponential is
adapted straight from 
[`def_disexp.R`](https://github.com/csgillespie/poweRlaw/blob/master/pkg/R/def_disexp.R#L78-L96).
These versions are also hard-coded for `log=TRUE`, the value required for
`compare_distributions`.


```{r}
dpldis2 <- function(x, xmin, alpha) {
    x <- x[round(x) >= round(xmin)]
    xmin <- floor(xmin)
    constant <- VGAM::zeta(alpha)
    if(xmin > 1) 
        constant <- constant - sum ((1:(xmin - 1)) ^ (-alpha))

    -alpha * log (x) - log (constant)
}
expdis <- function (m, q=NULL) 
{
    xmin <- m$getXmin()
    pars <- m$getPars()
    l1 <- pexp (q - 0.5, pars, lower.tail=FALSE, log.p=TRUE)
    l2 <- pexp (q + 0.5, pars, lower.tail=FALSE, log.p=TRUE)

    l1 + log (1 - exp (l2 - l1)) - 
        pexp (xmin - 0.5, pars, lower.tail=FALSE, log.p=TRUE)
}
compare_distributions2 <- function(d1, d2) 
{
    xmin <- d1$getXmin()
    q <- d1$getDat()
    q <- q[q >= xmin]
    ll_ratio_pts <- dpldis2 (q, xmin, m1$getPars()) - expdis(d2, q)

    m <- mean(ll_ratio_pts); s <- sd(ll_ratio_pts)
    v <- sqrt(length(ll_ratio_pts))*m/s
    p1 <- pnorm(v)

    if (p1 < 0.5) {p2 <- 2*p1} else {p2 <- 2*(1-p1)}

    l <- list(test_statistic = v, 
              p_one_sided = 1 - p1, p_two_sided=p2, 
              ratio = data.frame(x=q, ratio=ll_ratio_pts))
    class(l) <- "compare_distributions"
    return(l)
}
```
Then demonstrate that the above values are reproduced using these definitions:
```{r comp-dists2, eval = FALSE}
comp <- compare_distributions2 (m1, m2)
message ("test_statistic = ", comp$test_statistic, " (p = ", comp$p_two_sided, ")")
```

